---
title: "Exercise: ANOVA & Correlation"
---

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(knitr)
happy  <- read.csv("data/world_happiness.csv")
fifa   <- read.csv("data/fifa_players.csv")
movies <- read.csv("data/box_office.csv")
majors <- read.csv("data/college_majors.csv")
peek <- function(d, n = 5) kable(head(d, n))
```

## What you will do

1.  **Pick one** of the four datasets below.
2.  Run a **one-way ANOVA** (a numeric outcome across a categorical group). If the result is significant, run a **Tukey test** to see *which* groups differ.
3.  Run a **correlation analysis** on two numeric variables.

## How to load a dataset

```{r}
#| echo: true
#| eval: false
library(tidyverse)
library(mosaic)

read.csv("url of the data") %>%
  as_tibble() -> data
```

# Pick one dataset

## 1. World Happiness {#happiness}

The [World Happiness Report](https://worldhappiness.report/) scores countries on how happy people say they are, and estimates how much different factors contribute.

| Variable | Type | Meaning |
|---|---|---|
| `country` | non numeric | country name |
| `region` | categorical | Europe, Americas, Asia, Africa & Middle East |
| `happiness_score` | numeric | average happiness ladder score (0–10) |
| `gdp_per_capita` | numeric | the report's GDP-per-capita score (higher = richer) |
| `social_support` | numeric | having someone to count on |
| `life_expectancy` | numeric | healthy-life-expectancy score |
| `freedom` | numeric | freedom to make life choices |
| `generosity` | numeric | generosity score |

```{r}
#| echo: false
peek(happy)
```

-   **ANOVA:** does average `happiness_score` differ across `region`s?
-   **Correlation:** is `gdp_per_capita` correlated with `happiness_score`? (Does money buy happiness?)
-   **Also try:** is `generosity` correlated with `happiness_score`?

## 2. FIFA Football Players {#fifa}

Ratings and market values of footballers from the FIFA 19 database. Think of a player's market value as the **price of talent in a labour market**.

| Variable | Type | Meaning |
|---|---|---|
| `name`, `nationality`, `club` | non numeric | who the player is |
| `position` | categorical | Goalkeeper, Defender, Midfielder, Forward |
| `reputation` | categorical | international fame: `1-star` (least) … `3-star` (most) |
| `age` | numeric | age in years |
| `age_group` | categorical | age band: 21 or under, 22–25, 26–29, 30+ |
| `overall` | numeric | overall skill rating (0–100) |
| `potential` | numeric | potential rating |
| `value_million_eur` | numeric | market value, in million € |
| `wage_thousand_eur` | numeric | weekly wage, in thousand € |

```{r}
#| echo: false
peek(fifa)
```

-   **ANOVA:** does `value_million_eur` differ across `age_group`s? (Do players peak in market value at a certain age?)
-   **Correlation:** is `overall` rating correlated with `value_million_eur`? (Does skill set the price?)
-   **Also try:** does `value_million_eur` differ across `position`s? (Are forwards worth more than defenders — or not?)

## 3. Box Office {#boxoffice}

Budgets, ratings and worldwide box-office takings for Hollywood films. Think of a film as an investment.

| Variable | Type | Meaning |
|---|---|---|
| `title` | non numeric | film title |
| `genre` | categorical | Action, Adventure, Comedy, Drama, Horror |
| `budget_million` | numeric | production budget, in million US$ |
| `worldwide_gross_million` | numeric | worldwide box-office, in million US$ |
| `imdb_rating` | numeric | IMDb rating (0–10) |
| `rotten_tomatoes` | numeric | Rotten Tomatoes score (0–100) |
| `running_time_min` | numeric | length in minutes |
| `mpaa_rating` | non numeric | MPAA rating (G, PG, PG-13, R, …) |

```{r}
#| echo: false
peek(movies)
```

-   **ANOVA:** does `worldwide_gross_million` differ across `genre`s? (Do some genres earn more?)
-   **Correlation:** is `budget_million` correlated with `worldwide_gross_million`? (Does spending more earn more?)
-   **Also try:** is `imdb_rating` correlated with `worldwide_gross_million`? (Do better-rated films earn more?)

## 4. College Majors & Earnings {#majors}

Recent US graduates by major: what they earn, how many are women, and how often they're unemployed (US Census ACS data).

| Variable | Type | Meaning |
|---|---|---|
| `major` | non numeric | name of the major |
| `field` | categorical | Engineering & Computing, Sciences & Health, Business & Law, Social Sciences, Arts & Humanities |
| `median_earnings` | numeric | median earnings of recent graduates (US$) |
| `share_women` | numeric | share of graduates who are women (0–1) |
| `unemployment_rate` | numeric | unemployment rate (0–1) |
| `grads_total` | numeric | number of graduates |

```{r}
#| echo: false
peek(majors)
```

-   **ANOVA:** do `median_earnings` differ across `field`s of study? (Which fields pay most?)
-   **Correlation:** is `share_women` correlated with `median_earnings`? (Do fields with more women pay less?)
-   **Also try:** is `unemployment_rate` correlated with `median_earnings`?
