---
title: "One-Way ANOVA in R"
format:
  revealjs:
    slide-number: true
    incremental: true
    css: styles.css
    code-overflow: scroll
    code-annotations: hover
    footer: "**Anova**"
chunk_output_type: console
editor_options:
  chunk_output_type: console
---

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(mosaic)
```

# One-way ANOVA

## From t-tests to ANOVA {.small-slide}

-   In **Week 7** we used a [two-sample t-test]{.red-text} to compare the means of **two groups**.

-   But many real questions have more than two groups.

::: fragment
> "Are sales different across Monday, Tuesday, Wednesday, Thursday, and Friday?"
:::

::: fragment
> "Are sales different across front, back, left-aisle, and right-aisle locations?"
:::

::: fragment
> "Do students using different study methods have different exam scores?"
:::

::: fragment
> "Do patients receiving different treatments recover at different speeds?"
:::

::: fragment
For questions like these we use [one-way ANOVA]{.red-text}.
:::

## Why not many t-tests? {.small-slide}

-   If we have 4 locations, we could compare:

    -   front vs back
    -   front vs left-aisle
    -   front vs right-aisle
    -   back vs left-aisle
    -   back vs right-aisle
    -   left-aisle vs right-aisle

-   That is already **6 tests**.

::: fragment
The more tests we run, the more likely we are to find a "significant" result just by chance.
:::

::: fragment
ANOVA first asks one bigger question:

> "Is there evidence that at least one group mean is different?"
:::

# Our data

## Screen protector sales {.small-slide}

-   Imagine we work for an electronics store.

-   The store sells [screen protectors]{.red-text} for phones.

-   The manager wants to know what affects daily sales.

-   We have five variables:

    -   `weekday`: which day of the week
    -   `location`: where the screen protectors were placed
    -   `store_visitors`: number of customers who visited the store
    -   `discount_percent`: discount applied to screen protectors
    -   `units_sold`: how many screen protectors were sold

::: fragment
Today we will ask two different ANOVA questions using the same data.
:::

## Install the package {.small-slide}

::: fragment
Today we will use the `mosaic` package for easy summary statistics.
:::

::: fragment
```{r}
#| echo: true
#| eval: false
install.packages("mosaic") # <1>
```

1.  Run this only if you have not installed `mosaic` before. Installing a package once is enough.
:::

## Read the data {.small-slide}

```{r}
#| echo: true
#| eval: false
#| message: false
#| code-line-numbers: "1-2|3-4|5"
library(tidyverse) # <1>
library(mosaic) # <1>
read.csv("https://raw.githubusercontent.com/mucahitzor/IKT2010/refs/heads/main/data/screen_protector_sales.csv") %>% # <2>
  as_tibble() -> screen_sales # <3>
```

1.  Use `tidyverse` for data and plotting, and `mosaic` for easy summary statistics.
2.  Read the CSV file from GitHub.
3.  Save it in R's memory with the name `screen_sales`.

```{r}
#| echo: false
#| message: false
read.csv("data/screen_protector_sales.csv") %>%
  as_tibble() -> screen_sales
```

::: fragment
```{r}
#| echo: true
screen_sales
```
:::

::: fragment
Each row is one sales observation.
:::

::: fragment
-   `weekday`: which day of the week
-   `location`: where the screen protectors were placed
-   `store_visitors`: number of customers who visited the store
-   `discount_percent`: discount applied to screen protectors
-   `units_sold`: how many screen protectors were sold
:::

## Two questions today {.small-slide}

### Question 1

Do average screen protector sales differ by **weekday**?

::: fragment
$$
H_0:\ \mu_{Mon} = \mu_{Tue} = \mu_{Wed} = \mu_{Thu} = \mu_{Fri}
$$
:::

::: fragment
$$
H_a:\ \text{At least one weekday mean is different.}
$$
:::

### Question 2

Do average screen protector sales differ by **location**?

::: fragment
$$
H_0:\ \mu_{front} = \mu_{back} = \mu_{left} = \mu_{right}
$$
:::

::: fragment
$$
H_a:\ \text{At least one location mean is different.}
$$
:::

# Sales by weekday

## Mean sales by weekday {.small-slide}

::: fragment
Before doing any test, we first look at the group means.
:::

::: fragment
We use `favstats()` from the `mosaic` package.
:::

::: fragment
```{r}
#| echo: true
favstats(units_sold ~ weekday, data = screen_sales) # <1>
```

1.  Take `units_sold`, split it by `weekday`, and use the data set named `screen_sales`.

-   The `mean` column shows average sales for each weekday.

-   The weekday means are all around **24 to 26** units.

-   The `sd` column shows that sales vary quite a lot inside each weekday.
:::

::: fragment
So visually we expect a lot of overlap.

But we should not decide only from the means. Next we draw a boxplot.
:::

## Boxplot by weekday {.small-slide}

::: fragment
For a boxplot we need:

-   a group variable on the `x` axis
-   a numeric variable on the `y` axis
:::

::: fragment
```{r}
#| echo: true
screen_sales %>%
  ggplot() +
  aes(x = weekday, y = units_sold) +
  geom_boxplot()
```
:::

## Add fill color {.small-slide}

::: fragment
Now we can color the boxes by `weekday`.
:::

::: fragment
Because `weekday` is a variable, we put `fill = weekday` inside `aes()`.
:::

::: fragment
```{r}
#| echo: true
screen_sales %>%
  ggplot() +
  aes(x = weekday, y = units_sold, fill = weekday) +
  geom_boxplot()
```
:::

## Add a theme {.small-slide}

::: fragment
Finally we clean up the plot with a theme.
:::

::: fragment
```{r}
#| echo: true
screen_sales %>%
  ggplot() +
  aes(x = weekday, y = units_sold, fill = weekday) +
  geom_boxplot() +
  theme_minimal()
```
:::

## Interpreting the boxplot {.small-slide}

-   The boxes are at similar heights.

-   The medians are close to each other.

-   The boxes overlap a lot.

::: fragment
This suggests that weekday probably does **not** create a large difference in sales.
:::

::: fragment
But the boxplot hides the individual observations. Let's add the points.
:::

## Add the observations with geom_point() {.small-slide}

::: fragment
```{r}
#| echo: true
screen_sales %>%
  ggplot() +
  aes(x = weekday, y = units_sold, fill = weekday) +
  geom_boxplot() +
  geom_point() +
  theme_minimal()
```
:::

## Why geom_point() is not ideal {.small-slide}

-   The points are placed exactly on top of the weekday name.

-   If two observations have the same or similar sales value, they overlap.

-   This makes it hard to see how many observations we have.

::: fragment
We need to move the points slightly left and right.
:::

::: fragment
That is what `geom_jitter()` does.
:::

## Add the observations with geom_jitter() {.small-slide}

::: fragment
```{r}
#| echo: true
screen_sales %>%
  ggplot() +
  aes(x = weekday, y = units_sold, fill = weekday) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.15, alpha = 0.6) +
  theme_minimal()
```
:::

::: fragment
Now the points are easier to see, but each point still belongs to its weekday.
:::

## One-way ANOVA for weekday {.small-slide}

::: fragment
Now we ask the formal question:

> Are the weekday means different enough to reject $H_0$?
:::

::: fragment
```{r}
#| echo: true
aov(units_sold ~ weekday, data = screen_sales) # <1>
```

1.  Run ANOVA. The formula means: explain `units_sold` using `weekday`, with the data set named `screen_sales`.
:::

## Show the ANOVA table {.small-slide}

::: fragment
The previous code created the ANOVA model.
:::

::: fragment
Now we add `summary()` to see the ANOVA table.
:::

::: fragment
```{r}
#| echo: true
aov(units_sold ~ weekday, data = screen_sales) %>% # <1>
  summary() # <2>
```

1.  Run the same ANOVA model.
2.  Show the ANOVA table.
:::

## Reading the ANOVA table {.small-slide}

```{r}
#| echo: true
#| eval: false
aov(units_sold ~ weekday, data = screen_sales) %>%
  summary()
```

```{.text}
             Df Sum Sq Mean Sq F value Pr(>F)
weekday       4     63   15.86   0.272  0.895
Residuals   115   6699   58.25
```

The important line is the `weekday` line.

::: fragment
-   Let $k$ = number of groups, $N$ = total observations.
-   `Df`: $df_A = k - 1$ for `weekday`; $df_W = N - k$ for `Residuals`.
-   `Sum Sq`: `weekday` row is $SSA = \sum_j n_j(\bar{x}_j - \bar{x})^2$.
-   `Residuals` row is $SSW = \sum_j\sum_i (x_{ij} - \bar{x}_j)^2$.
-   Total variation is $SST = \sum_j\sum_i (x_{ij} - \bar{x})^2 = SSA + SSW$.
-   `Mean Sq`: $MSA = SSA / df_A$ and $MSW = SSW / df_W$.
-   `F value`: $F = MSA / MSW$.
-   `Pr(>F)`: the [p-value]{.red-text}, $P(F_{df_A,df_W} \geq F_{\text{observed}})$ if $H_0$ is true.
:::

::: fragment
For weekday, the p-value is about **0.895**.
:::

## Weekday decision {.small-slide}

-   Our decision rule:

    -   if $p < \alpha = 0.05$, reject $H_0$
    -   if $p \geq \alpha = 0.05$, fail to reject $H_0$

::: fragment
Here:

$$
p = 0.895 \geq \alpha = 0.05
$$
:::

::: fragment
So we [fail to reject $H_0$]{.red-text}.
:::

::: fragment
Conclusion:

> In this dataset, we do not have evidence that average screen protector sales differ by weekday.
:::

# Sales by location

## Mean sales by location {.small-slide}

::: fragment
Now we ask a different question with the same outcome.
:::

::: fragment
Instead of grouping sales by `weekday`, we group sales by `location`.
:::

::: fragment
```{r}
#| echo: true
favstats(units_sold ~ location, data = screen_sales) # <1>
```

1.  Take `units_sold`, split it by `location`, and use the data set named `screen_sales`.

-   The `front` location has the highest mean sales.

-   The `back` location has the lowest mean sales.

-   `left-aisle` and `right-aisle` are in the middle.
:::

::: fragment
This looks much more different than the weekday table.
:::

::: fragment
Now we check the distribution with a boxplot.
:::

## Boxplot by location {.small-slide}

::: fragment
```{r}
#| echo: true
screen_sales %>%
  ggplot() +
  aes(x = location, y = units_sold, fill = location) +
  geom_boxplot() +
  theme_minimal()
```
:::

## Interpreting the boxplot {.small-slide}

-   The `front` box is clearly higher.

-   The `back` box is clearly lower.

-   The aisle locations are between them.

::: fragment
The boxes overlap much less than the weekday boxes.
:::

::: fragment
This suggests that location may matter for screen protector sales.
:::

## Add the observations with geom_point() {.small-slide}

::: fragment
```{r}
#| echo: true
screen_sales %>%
  ggplot() +
  aes(x = location, y = units_sold, fill = location) +
  geom_boxplot() +
  geom_point() +
  theme_minimal()
```
:::

## Again, points overlap {.small-slide}

-   `geom_point()` puts all observations for the same location on one vertical line.

-   Some points hide other points.

-   We can see the general pattern, but the raw data are still crowded.

::: fragment
So again we use `geom_jitter()`.
:::

## Add the observations with geom_jitter() {.small-slide}

::: fragment
```{r}
#| echo: true
screen_sales %>%
  ggplot() +
  aes(x = location, y = units_sold, fill = location) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.15, alpha = 0.6) +
  theme_minimal()
```
:::

::: fragment
Now we see both the summary shape and the individual observations.
:::

## One-way ANOVA for location {.small-slide}

::: fragment
Now the formal question:

> Are the location means different enough to reject $H_0$?
:::

::: fragment
```{r}
#| echo: true
aov(units_sold ~ location, data = screen_sales) # <1>
```

1.  Run ANOVA. The formula means: explain `units_sold` using `location`, with the data set named `screen_sales`.
:::

## Show the ANOVA table {.small-slide}

::: fragment
Again, the first ANOVA code creates the model.
:::

::: fragment
Now we add `summary()` to see the ANOVA table.
:::

::: fragment
```{r}
#| echo: true
aov(units_sold ~ location, data = screen_sales) %>% # <1>
  summary() # <2>
```

1.  Run the same ANOVA model.
2.  Show the ANOVA table.
:::

## Reading the ANOVA table {.small-slide}

```{r}
#| echo: true
#| eval: false
aov(units_sold ~ location, data = screen_sales) %>%
  summary()
```

```{.text}
             Df Sum Sq Mean Sq F value Pr(>F)    
location      3   4845  1614.9   97.71 <2e-16 ***
Residuals   116   1917    16.5
```

The important line is the `location` line.

::: fragment
-   Let $k$ = number of groups, $N$ = total observations.
-   `Df`: $df_A = k - 1$ for `location`; $df_W = N - k$ for `Residuals`.
-   `Sum Sq`: `location` row is $SSA = \sum_j n_j(\bar{x}_j - \bar{x})^2$.
-   `Residuals` row is $SSW = \sum_j\sum_i (x_{ij} - \bar{x}_j)^2$.
-   Total variation is $SST = \sum_j\sum_i (x_{ij} - \bar{x})^2 = SSA + SSW$.
-   `Mean Sq`: $MSA = SSA / df_A$ and $MSW = SSW / df_W$.
-   `F value`: $F = MSA / MSW$.
-   `Pr(>F)`: the [p-value]{.red-text}, $P(F_{df_A,df_W} \geq F_{\text{observed}})$ if $H_0$ is true.
:::

::: fragment
For location, the p-value is very small: R reports **p < 2e-16**.
:::

## Location decision {.small-slide}

-   For `location`, the p-value is very small: R reports `p < 2e-16`.

::: fragment
That means:

$$
p < 0.05
$$
:::

::: fragment
So we [reject $H_0$]{.red-text}.
:::

::: fragment
Conclusion:

> In this dataset, average screen protector sales differ by location.
:::

# Which locations differ?

## ANOVA does not tell us where {.small-slide}

-   ANOVA answers one big question:

::: fragment
> "Is at least one group mean different?"
:::

-   But after a significant ANOVA, we usually want to know:

::: fragment
> "Which groups are different from which?"
:::

::: fragment
For that we use a [Tukey test]{.red-text}.
:::

## Tukey test for location {.small-slide}

::: fragment
```{r}
#| echo: true
aov(units_sold ~ location, data = screen_sales) %>% # <1>
  TukeyHSD() # <2>
```

1.  Run the same ANOVA model again.
2.  Compare all pairs of locations, while correcting for the fact that we are doing many comparisons.
:::

## Reading Tukey output {.small-slide}

```{.text}
$location
                             diff         lwr       upr     p adj
front-back              17.866667  15.1305127 20.602821 0.0000000
left-aisle-back          7.700000   4.9638461 10.436154 0.0000000
right-aisle-back         9.600000   6.8638461 12.336154 0.0000000
left-aisle-front       -10.166667 -12.9028206 -7.430513 0.0000000
right-aisle-front       -8.266667 -11.0028206 -5.530513 0.0000000
right-aisle-left-aisle   1.900000  -0.8361539  4.636154 0.2738489
```

-   `diff`: difference between two group means.

-   `lwr` and `upr`: lower and upper ends of the confidence interval.

-   `p adj`: adjusted p-value.

::: fragment
If `p adj < 0.05`, that pair is significantly different.
:::

::: fragment
If the confidence interval contains 0, that pair is not clearly different.
:::

## Visualize Tukey results {.small-slide}

::: fragment
The Tukey output is useful, but it has many numbers.
:::

::: fragment
We can also ask R to draw the Tukey confidence intervals.
:::

::: fragment
```{r}
#| echo: true
aov(units_sold ~ location, data = screen_sales) %>%
  TukeyHSD() %>%
  plot()
```
:::

::: fragment
If an interval crosses 0, that pair is not clearly different. If it does not cross 0, that pair is significantly different.
:::

## Tukey plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6.2
#| out-width: "100%"
old_par <- par(mar = c(5, 13, 4, 2))
aov(units_sold ~ location, data = screen_sales) %>%
  TukeyHSD() %>%
  plot(las = 1, cex.axis = 0.75)
par(old_par)
```

## Tukey result {.small-slide}

-   `front` is significantly higher than `back`.

-   `front` is significantly higher than `left-aisle`.

-   `front` is significantly higher than `right-aisle`.

-   `back` is significantly lower than the aisle locations.

::: fragment
But `left-aisle` and `right-aisle` are **not** significantly different from each other.
:::

::: fragment
Business interpretation:

> Put screen protectors in the front if you want higher sales. Left vs right aisle does not seem to matter much.
:::

# Summary

## What to remember {.small-slide}

-   A [one-way ANOVA]{.red-text} compares the means of **more than two groups**.

-   The outcome must be numeric: here, `units_sold`.

-   The group variable must be categorical: here, `weekday` or `location`.

-   Start with summary statistics: `favstats(units_sold ~ group, data = ...)`.

-   Then visualize: `geom_boxplot()` and `geom_jitter()`.

-   Then test: `aov(y ~ group, data = data) %>% summary()`.

-   If ANOVA is significant, use `TukeyHSD()` and `plot()` to see which groups differ.

## What we found {.small-slide}

| Question | p-value | Decision | Interpretation |
|---|---:|---|---|
| Sales by weekday | 0.895 | Fail to reject $H_0$ | Weekday does not clearly change sales |
| Sales by location | $< 2e^{-16}$ | Reject $H_0$ | Location clearly changes sales |

::: fragment
Same dataset. Same outcome. Different grouping variable. Different conclusion.
:::

## Your turn {.small-slide}

Using `screen_sales`:

1.  Use a different numeric variable: `store_visitors`.
2.  Make a boxplot of `store_visitors` by `location`.
3.  Add jittered points.
4.  Run the ANOVA: `store_visitors ~ location`.
5.  Explain the p-value in one sentence.

::: fragment
Then try the same outcome by weekday:

```{r}
#| echo: true
#| eval: false
aov(store_visitors ~ weekday, data = screen_sales) %>%
  summary()
```
:::

::: fragment
Ask yourself: does `weekday` explain store visitors as clearly as `location`?
:::

::: fragment
The most important sentence:

> "We reject / fail to reject $H_0$ because the p-value is smaller / larger than 0.05."
:::

# Questions?

## See you next week
