---
title: "Scatter Plots & Relationships"
format:
  revealjs:
    slide-number: true
    incremental: true
    css: styles.css
    code-overflow: scroll
    code-annotations: hover
    footer: "**Week 4 - English**"
chunk_output_type: console
editor_options:
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(janitor)
library(ggthemes)
```

# What we learned so far

## Recap {.small-slide}

-   We learned how to visualize the [distribution of one continuous variable]{.red-text}.

-   **Histogram**: groups data into bins, counts how many observations fall in each bin.

-   **Density plot**: estimates a smooth curve representing the distribution.

-   We also learned how to compare distributions across groups (e.g., income by gender).

::: fragment
But what if we want to ask a different kind of question?
:::

::: fragment
> "Is there a **relationship** between two variables?"
:::

## New Question

-   In Week 3 we asked: *"How is income distributed?"* → one variable.

-   This week we ask: *"Does income inequality relate to hate crime rates?"* → [two variables]{.red-text}.

-   When we want to see the **relationship** between two continuous variables, we use a **scatter plot**.

## When to use a scatter plot?

-   You have **two continuous variables**.

-   You want to see if there is a **pattern**, **trend**, or **correlation** between them.

-   Each observation (row) becomes **one point** on the plot.

-   The `x` axis shows one variable, the `y` axis shows the other.

# Our Data: Hate Crimes in the US

## The Data {.small-slide}

-   Our data comes from a [FiveThirtyEight](https://fivethirtyeight.com/) article:

    -   *"Higher Rates Of Hate Crimes Are Tied To Income Inequality"*

-   It contains data on all 50 US states.

-   For each state we have:

    -   `hate_crimes_fbi`: hate crimes per 100,000 people (reported to FBI)
    -   `gini_index`: a measure of income inequality (higher = more unequal)
    -   `median_income`: median household income
    -   `highschool`: share of population with a high school degree
    -   `unemployment`: share of unemployed population
    -   `trump_vote`: share of voters who voted for Trump in 2016
    -   `region`: US Census region (Northeast, Midwest, South, West)

## Read the data {.small-slide}

```{r warning=FALSE}
#| echo: true
#| eval: true
#| message: false
#| code-line-numbers: "1-3|4-5"
library(tidyverse) # <1>
library(janitor) # <1>
library(ggthemes) # <1>
read.csv("data/hate_crimes.csv") %>%  # <2>
  as_tibble() -> hate_crimes # <3>
```

1.  Use the tidyverse, janitor, and ggthemes libraries
2.  Read the CSV file and then do the following ( %>% )
3.  Show it as a `tibble` and save it as `hate_crimes`

## Have a look at the data

::: fragment
```{r}
#| echo: true
hate_crimes
```
:::

## Understanding Our Data {.small-slide}

### What are our variables?

::: fragment
```{r}
#| echo: true
hate_crimes %>%
  summary()
```
:::

## Understanding Our Data {.small-slide}

### Non-numeric variable: region

::: fragment
```{r}
#| echo: true
hate_crimes %>%
  tabyl(region)
```
:::

-   We have 4 regions, each with a different number of states.

## Our Goal Today {.small-slide}

### Is there a relationship between income inequality and hate crimes?

-   `gini_index` measures income inequality. Higher values = more inequality.

-   `hate_crimes_fbi` measures hate crimes per 100,000 people.

-   We want to see: do states with **more inequality** also have **more hate crimes**?

-   Both variables are **continuous** → we need a **scatter plot**.

# Building a Scatter Plot Step by Step

## Step 1: Start with the data {.small-slide}

::: fragment
Take our data set named `hate_crimes`,
:::

::: fragment
```{r}
#| echo: true
#| eval: false
hate_crimes
```
:::

::: fragment
I want to make a visualization, take my data set and apply `ggplot`.
:::

::: fragment
```{r}
#| echo: true
hate_crimes %>%
  ggplot()
```
:::

## Step 2: Set the axes {.small-slide}

::: fragment
Now we need to tell R what goes on the `x` axis and what goes on the `y` axis.
:::

::: fragment
For scatter plots we need **both** `x` and `y` inside `aes()`.
:::

::: fragment
-   `x = gini_index` (income inequality)
-   `y = hate_crimes_fbi` (hate crimes per 100k)
:::

::: fragment
```{r}
#| echo: true
hate_crimes %>%
  ggplot() +
  aes(x = gini_index, y = hate_crimes_fbi)
```
:::

## Step 3: Add the points {.small-slide}

::: fragment
Remember, for histograms we used `geom_histogram()`, for density we used `geom_density()`.
:::

::: fragment
For scatter plots we use `geom_point()`.
:::

::: fragment
Each row in our data becomes **one point** on the plot.
:::

::: fragment
```{r}
#| echo: true
hate_crimes %>%
  ggplot() +
  aes(x = gini_index, y = hate_crimes_fbi) +
  geom_point()
```
:::

## Step 4: Add a theme {.small-slide}

::: fragment
Lets immediately clean it up with a theme.
:::

::: fragment
```{r}
#| echo: true
hate_crimes %>%
  ggplot() +
  aes(x = gini_index, y = hate_crimes_fbi) +
  geom_point() +
  theme_fivethirtyeight()
```
:::

## Step 5: Customize the points {.small-slide}

::: fragment
The default points are small and hard to see. We can change them inside `geom_point()`:
:::

::: fragment
-   `size` = how big the points are (default is about 1.5)
-   `color` = the color of the points
:::

::: fragment
```{r}
#| echo: true
hate_crimes %>%
  ggplot() +
  aes(x = gini_index, y = hate_crimes_fbi) +
  geom_point(size = 3, color = "steelblue") +
  theme_fivethirtyeight()
```
:::

## Step 6: Fix the axes {.small-slide}

::: fragment
Like in Week 3, lets increase the number of values on both axes.
:::

::: fragment
```{r}
#| echo: true
hate_crimes %>%
  ggplot() +
  aes(x = gini_index, y = hate_crimes_fbi) +
  geom_point(size = 3, color = "steelblue") +
  scale_x_continuous(n.breaks = 10) +
  scale_y_continuous(n.breaks = 10) +
  theme_fivethirtyeight()
```
:::

## Interpreting the scatter plot {.small-slide}

-   What do we see?

    -   As `gini_index` **increases** (more inequality), hate crimes **tend to increase** too.
    -   This suggests a [positive relationship]{.red-text} between income inequality and hate crimes.

-   But not all states follow this pattern perfectly.

    -   Some states have high inequality but low hate crimes.
    -   Some states have low inequality but high hate crimes.
    -   This is normal — we are looking at a **trend**, not a perfect rule.

-   There is one point very high up — can you guess which state that might be?

::: fragment
```{r}
#| echo: true
hate_crimes %>% filter(hate_crimes_fbi > 10)
```
:::

## Add labels {.small-slide}

::: fragment
Now lets add a `title`, `subtitle`, and axis labels using `labs()`.
:::

::: fragment
```{r}
#| echo: true
#| eval: false
hate_crimes %>%
  ggplot() +
  aes(x = gini_index, y = hate_crimes_fbi) +
  geom_point(size = 3, color = "steelblue") +
  scale_x_continuous(n.breaks = 10) +
  scale_y_continuous(n.breaks = 10) +
  labs(
    title = "Income Inequality and Hate Crimes in US States",
    subtitle = "States with higher Gini index tend to have more hate crimes per 100k people",
    x = "Gini Index (Income Inequality)",
    y = "Hate Crimes per 100,000 (FBI)"
  ) +
  theme_fivethirtyeight()
```
:::

## Add labels {.small-slide}

::: fragment
```{r}
#| echo: false
hate_crimes %>%
  ggplot() +
  aes(x = gini_index, y = hate_crimes_fbi) +
  geom_point(size = 3, color = "steelblue") +
  scale_x_continuous(n.breaks = 10) +
  scale_y_continuous(n.breaks = 10) +
  labs(
    title = "Income Inequality and Hate Crimes in US States",
    subtitle = "States with higher Gini index tend to have more hate crimes per 100k people",
    x = "Gini Index (Income Inequality)",
    y = "Hate Crimes per 100,000 (FBI)"
  ) +
  theme_fivethirtyeight()
```
:::

# Coloring Points by a Group

## Color by region {.small-slide}

-   Remember in Week 3 we changed the `color` of density lines based on `gender`?

-   We can do the same thing here: color the points based on `region`.

-   Since `region` is a **variable**, we put it inside `aes()`, not inside `geom_point()`.

::: fragment
Can you guess the code?
:::

## Color by region {.small-slide}

::: fragment
We move `color` from `geom_point(color = "steelblue")` to `aes(color = region)`:
:::

::: fragment
```{r}
#| echo: true
hate_crimes %>%
  ggplot() +
  aes(x = gini_index, y = hate_crimes_fbi, color = region) +
  geom_point(size = 3) +
  scale_x_continuous(n.breaks = 10) +
  scale_y_continuous(n.breaks = 10) +
  labs(
    title = "Income Inequality and Hate Crimes in US States",
    subtitle = "Colored by US Census region",
    x = "Gini Index (Income Inequality)",
    y = "Hate Crimes per 100,000 (FBI)"
  ) +
  theme_fivethirtyeight()
```
:::

## What do we see? {.small-slide}

-   The **South** (red/pink) tends to cluster in the lower-right: high inequality, but lower FBI-reported hate crimes.

-   The **Northeast** tends to have higher hate crime rates.

-   Does this mean the South has fewer hate crimes? Not necessarily — it could mean different reporting practices across regions.

-   This is an important lesson: [correlation does not mean causation]{.red-text}, and data quality matters!

## Try it yourself! {.small-slide}

-   Now it is your turn. Using the same `hate_crimes` data:

    1.  Make a scatter plot of `median_income` (x axis) vs `hate_crimes_fbi` (y axis).
    2.  Color the points by `region`.
    3.  Add a nice theme.
    4.  Add a title and axis labels.
    5.  What pattern do you see? Do richer states have more or fewer hate crimes?

# Making Points Transparent

## The alpha argument {.small-slide}

-   Sometimes points overlap and we cannot see how many points are in the same area.

-   We can make points semi-transparent using `alpha` inside `geom_point()`.

-   `alpha = 1` means fully solid (default), `alpha = 0` means invisible.

::: fragment
```{r}
#| echo: true
hate_crimes %>%
  ggplot() +
  aes(x = gini_index, y = hate_crimes_fbi, color = region) +
  geom_point(size = 3, alpha = 0.7) +
  scale_x_continuous(n.breaks = 10) +
  scale_y_continuous(n.breaks = 10) +
  labs(
    title = "Income Inequality and Hate Crimes in US States",
    subtitle = "Colored by US Census region",
    x = "Gini Index (Income Inequality)",
    y = "Hate Crimes per 100,000 (FBI)"
  ) +
  theme_fivethirtyeight() +
  scale_color_fivethirtyeight()
```
:::

# Adding a Trend Line

## What is a trend line? {.small-slide}

-   A scatter plot shows us the individual data points.

-   But sometimes it is hard to see the **overall pattern** just from the points.

-   A **trend line** (also called a regression line or line of best fit) helps us see the general direction of the relationship.

-   If the line goes **up** → positive relationship (as x increases, y increases).

-   If the line goes **down** → negative relationship (as x increases, y decreases).

-   If the line is **flat** → no clear relationship.

## Adding a trend line with geom_smooth() {.small-slide}

::: fragment
We use `geom_smooth()` to add a trend line on top of our scatter plot.
:::

::: fragment
We add `method = "lm"` inside `geom_smooth()` to get a straight line. `lm` stands for **linear model**.
:::

::: fragment
```{r}
#| echo: true
#| message: false
hate_crimes %>%
  ggplot() +
  aes(x = gini_index, y = hate_crimes_fbi) +
  geom_point(size = 3, color = "steelblue") +
  geom_smooth(method = "lm") +
  theme_fivethirtyeight()
```
:::

## Understanding the trend line {.small-slide}

-   The blue line shows the **overall trend**: as inequality increases, hate crimes tend to increase.

-   The **gray shaded area** around the line shows the **confidence interval** — it tells us how uncertain the trend is.

    -   A narrow band means we are more confident.
    -   A wide band means we are less confident.

-   We can remove the confidence interval with `se = FALSE`:

::: fragment
```{r}
#| echo: true
#| message: false
hate_crimes %>%
  ggplot() +
  aes(x = gini_index, y = hate_crimes_fbi) +
  geom_point(size = 3, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  theme_fivethirtyeight()
```
:::

## Putting it all together {.small-slide}

::: fragment
```{r}
#| echo: true
#| eval: false
#| message: false
hate_crimes %>%
  ggplot() +
  aes(x = gini_index, y = hate_crimes_fbi) +
  geom_point(size = 3, color = "steelblue", alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  scale_x_continuous(n.breaks = 10) +
  scale_y_continuous(n.breaks = 10) +
  labs(
    title = "Income Inequality and Hate Crimes in US States",
    subtitle = "A positive trend: states with higher inequality tend to report more hate crimes",
    x = "Gini Index (Income Inequality)",
    y = "Hate Crimes per 100,000 (FBI)"
  ) +
  theme_fivethirtyeight()
```
:::

## Putting it all together {.small-slide}

::: fragment
```{r}
#| echo: false
#| message: false
hate_crimes %>%
  ggplot() +
  aes(x = gini_index, y = hate_crimes_fbi) +
  geom_point(size = 3, color = "steelblue", alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  scale_x_continuous(n.breaks = 10) +
  scale_y_continuous(n.breaks = 10) +
  labs(
    title = "Income Inequality and Hate Crimes in US States",
    subtitle = "A positive trend: states with higher inequality tend to report more hate crimes",
    x = "Gini Index (Income Inequality)",
    y = "Hate Crimes per 100,000 (FBI)"
  ) +
  theme_fivethirtyeight()
```
:::

## Trend line by group {.small-slide}

::: fragment
What happens if we add `color = region` back into `aes()`?
:::

::: fragment
```{r}
#| echo: true
#| message: false
hate_crimes %>%
  ggplot() +
  aes(x = gini_index, y = hate_crimes_fbi, color = region) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Income Inequality and Hate Crimes by Region",
    subtitle = "Each region gets its own trend line",
    x = "Gini Index (Income Inequality)",
    y = "Hate Crimes per 100,000 (FBI)"
  ) +
  theme_fivethirtyeight() +
  scale_color_fivethirtyeight()
```
:::

## Trend line by group {.small-slide}

-   When `color = region` is inside `aes()`, **both** `geom_point()` and `geom_smooth()` split by region.

-   Each region gets its own points and its own trend line.

-   This is useful to see if the relationship is different across groups.

-   The relationship direction varies by region — another reason to be careful about making conclusions!

# Summary

## What we learned today {.small-slide}

-   **Scatter plot**: shows the relationship between two continuous variables using `geom_point()`.

-   **Building a scatter plot**:
    1.  `ggplot()` → blank canvas
    2.  `aes(x = ..., y = ...)` → set both axes
    3.  `geom_point()` → add the points
    4.  Customize with `size`, `color`, `alpha`
    5.  `labs()` → add title, subtitle, axis labels

-   **Color by group**: put `color = variable` inside `aes()` to color points by a categorical variable.

-   **Trend line**: `geom_smooth(method = "lm")` adds a line of best fit.

-   **Key takeaway**: [Correlation does not imply causation!]{.red-text} Just because two variables move together does not mean one causes the other.

## Exercise {.small-slide}

Using the `hate_crimes` data, explore one of these questions:

1.  Is there a relationship between `unemployment` and `hate_crimes_fbi`?
2.  Is there a relationship between `highschool` (education) and `hate_crimes_fbi`?
3.  Is there a relationship between `median_income` and `gini_index`?

For each:

-   Make a scatter plot with `geom_point()`
-   Add a trend line with `geom_smooth(method = "lm")`
-   Color by `region`
-   Add a title and axis labels
-   Interpret: what does the plot tell you?
